Capabilities, MMIO and Robust Safety

### 2020-03-24 capability-meetings-notes/10-03/main.tex
MMIO meeting notes: a few improvements / cleanups

\newcommand{\MMIO}{\textlog{MMIO}}

\DeclareMathOperator{\initOKo}{init_{OK}}
\newcommandx{\initOK}[4][3=r,4=m]{\initOKo(#1,#2,#3,#4)}
\DeclareMathOperator{\initBCo}{init_{BC}}
\newcommandx{\initBC}[4][3=r,4=m]{\initBCo(#1,#2,#3,#4)}

\newcommand{\regdr}{\mathbin{!!}}

@@

\textit{NB:} An alternative presentation would be to make this range an
immutable part of \X{State}. It probably does not really matter in the end, we
should do whatever makes our life easier for instantiating Iris with the
semantics.

@@

With MMIO, we obtain two rules:

@@

  \inferrule[IOLoad]
  { r[\X{src}] = (p,g,b,e,a) \\ \X{readAllowed}\; p \\ a \in [b,e) \\ a \in \MMIO}
  { (r, m, t) \xrightarrow{\I{Load} \; \X{dst} \; \X{src}}
    (r[\X{dst} := x], m, (\X{IORead}, a, x) :: t) }
\end{mathpar}

@@

The \I{Store} rule would be similar. Additionally, executing with a \X{PC}
register pointing to an $\MMIO$ location should be disallowed in the operational
semantics (I think this makes more sense than reading a random instruction
through I/O).

\paragraph*{Remark:} With this presentation, the specification of the
``machine'' is very much decoupled from the model of the devices it might be
communicating with. This is a good thing, I think.

@@

I believe this would be somewhat similar to the semantics of I/O system calls
through a foreign function interfaces as formalized in
CakeML~\cite{cakeml-vstte17io}.

\paragraph*{Remark:} Alix says that the proposed style of operational semantics
(using a trace) is close to the semantics of \texttt{volatile} as in
CompCert---which is also how memory mapped addresses seem to be exposed to C
compilers in practice. So this is probably a good sign.

@@

We added a trace as part of the state, so we wish to also expose it as a \SL
assertion.

@@

The simplest way to do that is to directly expose the complete trace. In that
case, the state interpretation additionally holds a resource for the trace:

\[
  \X{state\_interp} \; (r,m,t) := \ldots \ast \tracefull{t}
\]

The user then works with assertions of the form $\tracefrag{t'}$. Such an
assertion is not duplicable, and grants full ownership over the trace. In
particular, it allows one to \emph{update} the trace by emitting events, i.e. by
performing I/O operations.

@@

\paragraph*{Remark:} This is very coarse-grained: either one has the full
ownership for performing I/O and reasoning about it, or one cannot know anything
about it.

@@

One application of this second extension could be the verification of an example
involving ``multiplexed'' I/O, where two separate parts of the code are granted
separate ownership over separate MMIO addresses. These two separate pieces of
code would be verified separately; then, in the end, one could prove that one
gets \emph{some} interleaving of all the events emitted by both components.

We use another resource algebra to model the MMIO locations. The operational
semantics have been parameterized with a set $\MMIO$ of memory locations, but we
need a way to specify what these locations are at a higher level of abstraction.
A non-elegant solution is to index the WP definition with this set $\MMIO$. A
better solution (that does not necessitate any changes to the definition of WP)
is to add $\MMIO$ to the state, and add a conjunct to the state interpretation
that expresses ownership of $\MMIO$ under agreement, i.e. add

\newcommand{\MMIOag}{\ownGhost{\gamma_{\X{MMIO}}}{\agm (\X{MMIO})}}

\[
  \X{state\_interp} \; (r,m,t,\MMIO) := \ldots \ast \MMIOag
\]

to the state interpretation we had before.

\paragraph*{Remark:} Another alternative, mentioned by Dominique, was to
parameterize the entire language by the set of MMIO locations. This might come
down to the same thing as the first solution mentioned above.

In the future, we might be interested in the dynamic allocation of MMIO memory.
The question is what this would mean, though, since the set of available devices
allowing for MMIO access and the concrete buffers they provide will most likely
still be modeled as fixed at runtime.
Rather, this would allow us to take (un)mapped MMIO locations, connected to the different devices, and (un)map them anywhere in main memory.
This would require a different way of modeling MMIO, where we need both a notion
of the total pool of possible MMIO locations, and a description of the currently
mapped locations. We could model this using an authoritative RA.

@@

One then wants to be able to encapsulate the use of $\OutV$, for instance by
defining object capabilities that enforce some invariants on the values being
printed. In that setting, $\OutV$ is considered as a ``high'' value:

\[
  \begin{array}{lcl}
    \liftP \OutV & \eqdef & \FALSE
  \end{array}
\]

To be more precise, we could actually allow sharing $\OutV$ in case the predicate P
does not place any constraints on the output, i.e. is the $\TRUE$ predicate.
Technically, we could hence have the following definition (although admittedly,
it does look a bit hard-coded):
\[
  \begin{array}{lcl}
    \liftP \OutV & \eqdef & \exists \iota \ldotp \knowInv{\iota}{\exists t\ldotp \tracefrag{t} \ast
                            \pure{(\Lam{\_}.\TRUE) \spac t}}
  \end{array}
\]

The theorem then becomes:

\begin{mathpar}
  \mprset{vskip=0.4em}
  \inferrule
  {C \in \textit{AdvCtx} \\
    e \; \X{closed} \\
    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast \pure{\! P(t)}} \vdash \hoare{\TRUE}{e}{x.\; \textlog{lowval}\; x} \\
    (C[e]);(\emptyset, \textlog{OK}); \emptyset \longrightarrow^* T';(h',g'); t
  }
  {g' = \textlog{OK} \wedge P(t)}
\end{mathpar}

@@

@@

@@

Then, $\VR^{\notMMIO}$ is similar to the $\textlog{lowval}$ predicate of OCPL.
Intuitively, a value in the relation does not directly point to memory-mapped
locations, nor can it gain access to memory-mapped locations indirectly,
similarly to how $\textlog{lowval}$s cannot grant access to high locations
(in)directly. The relation $\ER^{\notMMIO}$ specifies that, if we fill the
registers with values in $\VR^{\notMMIO}$, then the
invariants in a private future world of our starting world will be satisfied at the end of execution.

\paragraph*{Remark:} As stated above, the ``generalized'' value relation is in
fact not very useful for instantiation of $\Psi$ other than $\notMMIO$. Indeed,
even if $\Psi$ \emph{does} allow referring to addresses in the $\MMIO$ region,
the value relation does not grant any corresponding resources.
The fix would be to use the generalized trace resources mentioned previously,
and grant ownership for the part of the trace corresponding to the
addresses in $[b,e) \cap \MMIO_{\X{pub}}$ with $\MMIO_{\X{pub}}$ the shareable subset of addresses in \MMIO.

We will have to adapt the robust safety theorem to this new capability machine
setting, \'and to the more concrete setup of the examples we want to gradually
verify.

Ignoring this second restriction for the time being, the general updated robust
safety theorem might look something like this:

\begin{mathpar}
  \mprset{vskip=0.4em}
  \inferrule
  { \forall a_{i} \in A_{\X{entry}} \ldotp \left\{
      {\begin{array}{ll}
      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\! \ast \MMIOag
         \vdash\\
         \quad \forall W \ldotp\;(\ER^{\notMMIO} (RX, g, b, e, a_i))\spac W
      \end{array}} \right. \\
     \\
    (r[\overline{r_{a_{i}} := (E, g, b, e, a_i)}], m, \emptyset) \longrightarrow^*
    (r', m', t)
  }
  {P(t)}
\end{mathpar}

The capability $(RX, g, b, e, a_i)$ points to the $i$th entry point of the code
we want to gradually verify.
The set of entry points is given by $A_{\X{entry}}$.
The capability $(E, g, b, e, a_i)$ then represents
a closure for this same piece of code. The adversary only gets access to the
closure, in order to avoid them from executing or reading arbitrary lines of
code within the trusted module.
Notice the universal quantification over worlds in the precondition;
this needs to be there, since we do not know in what world our code will be
called.
This universal quantification was not explicitly present in the OCPL
formalization.
The reason for that is that (I think so at least, do not quote me
on this - Thomas) they
used the built-in Iris worlds within their definition of weakest precondition,
by leveraging the state interpretation, which is universally quantified over in
the definition of weakest precondition.
As we discussed before, if we do away
with local capabilities, it should be possible to fall back onto Iris' notion of
worlds, which would allow us to remove the universal quantification in the
above definition too because it would be implicitly present in the WP inside $\ER$.

\paragraph*{Remark:} it would be interesting to have a discussion about how to
do away with the worlds if we remove local capabilities, regardless of whether
we will in the end, since that might shed some light on why the worlds look the
way they do right now.

Also note that we might as well have written
$\ldots \vdash \forall W \ldotp \; (\VR^{\notMMIO} \ldots $ %)
in the above theorem, instead of having the expression relation, since this is
the actual relation we will need, and easily unfolds to the former statement
through the definition of the value relation. I wrote $\ER$ to place the focus
on the execution of the driver being safe. I consider this a matter of taste.

The above theorem is still lacking, since it has no notion of adversarial
context \textit{AdvCtx} to limit the power of the attacker. Indeed, if we allow
an attacker to have read or write access to any location in \MMIO, or possess
any capability directly pointing into the driver, then all hope we might have of
gradually enforcing safety properties is now lost. We need a check on the memory
and registers that we will allow our driver closures to be used with.

To enforce these syntactic restrictions, we have two options:
\begin{enumerate}
\item Add a syntactic check to the statement of our robust safety theorem,
  ensuring that the adversary code cannot gain access to any non-E capabilities
  pointing into the driver code or any capabilities pointing into $\MMIO$ directly.
\item Assume the existence of a piece of boot code, that gets to run first when
  the capability machine starts up, and makes sure the above condition actually
  holds before passing control to the adversary. We could either assume that the
  adversarial memory can contain any capability at startup, in which case the
  driver has to erase the memory before passing control, or we could assume that
  the memory cannot contain any capabilities whatsoever, in which case the boot
  code does not have this obligation.
\end{enumerate}

If we were to follow option 2 and write concrete boot code, part of our proof
would consist of reducing option 2 to the situation in option 1, since this is
where unknown adversary code actually starts executing. In a way, option 1 is
hence more fundamental than option 2. We therefore adapt the robust safety
theorem to option 1 first.

In this setting, we can subdivide the memory $m$ into 3 relevant subsections;
$m = m_{\MMIO} \uplus m_{\X{driver}} \uplus m_{\X{adv}}$ %\uplus m_{\X{frame}}
with the following meaning:
\begin{itemize}
\item $m_{\MMIO}$ contains the memory-mapped locations
\item $m_{\X{driver}}$ contains the driver code
\item  $m_{\X{adv}}$ contains the adversary's code and data (this will also
  include the boot code in the general setting, since it does not contain any
  inherently dangerous capabilities itself)
\end{itemize}.

@@

We can then implement the syntactic check using the following predicate
$\notMapsToR{R}{w}$, that takes a word, memory region or register file and checks if it points into the forbidden region $R$ as follows:

\[
  \begin{array}{lcl}
    \notMapsToR{R}{\X{inl}\spac z} & \eqdef
    & \TRUE \\
    %
    \notMapsToR{R}{\X{inr}\spac (p,g,b,e,a)} & \eqdef & R \cap [b,e[\; =
                                                 \emptyset  \\
    %
    \notMapsToR{R}{\V{reg}} & \eqdef & \forall r \in \dom(\V{reg}) \ldotp\spac \notMapsToR{R}{\V{reg}(r)} \\
    %
    \notMapsToR{R}{\V{mem}} & \eqdef & \forall l \in \dom(\V{mem}) \ldotp\spac   \notMapsToR{R}{\V{mem}(l)}
  \end{array}
\]

Given this definition, we can now formalize the initialization constraints on
registers $r$ and memory $m$, given the region $\MMIO$ and
the driver's address space $[b,e[$ as follows;

\begin{mathpar}
  \mprset{vskip=0.4em}
  \inferrule
  { m = m_{\MMIO} \uplus m_{\X{driver}} \uplus m_{\X{adv}}\\
    m_{\MMIO} = \MMIO \rightarrow \mbox{\Lightning} \\
    \dom( m_{\X{driver}} ) = [b,e[\\
    \dom( m_{\X{adv}} ) = [b_{\X{adv}},e_{\X{adv}}[\\
    \notMapsToR{\MMIO \cup [b,e[}{\V{r}}\\
    \notMapsToR{\MMIO \cup [b,e[}{\V{m_{\X{adv}}}} \\
    r \regdr \X{PC} = (\X{RWX}, \X{G},b_{\X{adv}},e_{\X{adv}},b_{\X{adv}})\\
  }
  {\initOK{\MMIO}{[b,e[}}
\end{mathpar}

@@

\begin{mathpar}
  \mprset{vskip=0.4em}
  \inferrule
  { \forall a_{i} \in A_{\X{entry}} \ldotp \left\{
      {\begin{array}{ll}
      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\! \ast \MMIOag
         \vdash\\
         \quad \forall W \ldotp\;(\ER^{\notMMIO} (RX, g, b, e, a_i))\spac W
       \end{array}} \right. \\
   \initOK{\MMIO}{[b,e[} \\
  (r[\overline{r_{a_{i}} := (E, g, b, e, a_i)}], m, \emptyset) \longrightarrow^* (r', m', t)\\
  }
  {P(t)}
\end{mathpar}
, where we have set the PC up to be a sensible capability (pointing to $b_{\X{adv}}$
without loss of generality).

@@

Having this theorem in our toolbox, we can now take a closer look at scenario 2
above, where a piece of boot code is used to set the memory up correctly,
obviating the need for any other assumptions than that the boot code behaves
according to some spec. Concretely, we want the boot code to uphold a spec that
allows us to prove the above theorem.
Alternatively, we could also just implement one specific instance of boot code, and prove that after it finishes executing, all preconditions to apply the above theorem are met.
We can use the following predicate to specify that the starting configuration is
valid:

\begin{mathpar}
  \mprset{vskip=0.4em}
  \inferrule
  { m = m_{\MMIO} \uplus m_{\X{driver}} \uplus m_{\X{boot}}\\
    m_{\MMIO} = \MMIO \rightarrow \mbox{\Lightning} \\
    \dom(m_{\X{driver}}) = [b,e[\\
    \dom(m) = [0,\X{MEM_{MAX}}[\\
    r \regdr \X{PC} = (\X{RWX}, \X{G},0,\X{MEM_{MAX}},l_{\X{boot}})\\
  }
  {\initBC{\MMIO}{[b,e[}}
\end{mathpar}
where $l_{\X{boot}}$ is the location where the machine boots, to be determined
by the machines operational semantics.

On the level of the operational semantics, we have to prove that the
following holds, in order to use our concrete driver implementation with the
robust safety theorem above (the driver still needs to be proven safe separately):

\begin{mathpar}
  \mprset{vskip=0.4em}
  \inferrule
  { \initBC{\MMIO}{[b,e[}\\
  }
  { \exists n\ldotp (r, m, \emptyset) \longrightarrow^n (r'[\overline{r_{a_{i}}
      := (E, g, b, e, a_i)}], m', t) \land P(t) \land\\
  \initOK{\MMIO}{[b,e[}[r'][m']}\quad (\textsc { BootCodeOK })
\end{mathpar}

Note that to link scenario 2 to scenario 1, we need a slightly adapted version of the robust safety theorem stated above
where the initial trace is allowed to not be empty, but rather just satisfy
$P(t)$, but that should not cause any problems.
For a specific instance of boot code that provably satisfies the \textsc{BootCodeOK}
rule, the resulting boot-code-based formulation of the robust safety theorem is now
just a corollary of the general robust safety theorem, and has the following
statement:

\begin{mathpar}
  \mprset{vskip=0.4em}
  \inferrule
  { \forall a_{i} \in A_{\X{entry}} \ldotp \left\{
      {\begin{array}{ll}
      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\! \ast \MMIOag
         \vdash\\
         \quad \forall W \ldotp\;(\ER^{\notMMIO} (RX, g, b, e, a_i))\spac W
       \end{array}} \right. \\
   \initBC{\MMIO}{[b,e[} \\
  (r, m, \emptyset) \longrightarrow^* (r', m', t)\\
  }
  {P(t)}
\end{mathpar}

Now we have a closer look at the concrete example we would like to verify,
including what its boot code looks like. Our goal is to prove that this concrete
example satisfies \textsc{BootCodeOk} and this last robust safety theorem.

>>>>

\newcommand{\MMIO}{\textlog{MMIO}\xspace}

\DeclareMathOperator{\initOKo}{init_{OK}}
\newcommandx{\initOK}[3][2=r,3=m]{\initOKo(#1,#2,#3)}
\DeclareMathOperator{\initBCo}{init_{BC}}
\newcommandx{\initBC}[2][1=r,2=m]{\initBCo(#1,#2)}

\newcommand{\app}{\mathbin{+\!\!+}}

\newenvironment{remark}
{ \bigskip\hrule\vspace{-1.3em}\nobreak
  \paragraph*{Remark:}}
{\vspace*{0.5em}\hrule\medskip}

@@

@@

With MMIO, we obtain two rules for \I{Load}:

@@

  \inferrule[IOLoad]
  { r[\X{src}] = (p,g,b,e,a) \\ \X{readAllowed}\; p \\ a \in [b,e) \\ a \in \MMIO}
  { (r, m, t) \xrightarrow{\I{Load} \; \X{dst} \; \X{src}}
    (r[\X{dst} := x], m, t \app (\X{IORead}, a, x)) }
\end{mathpar}

@@

The \I{Store} rule would be similar.

Notice that, for any address in \MMIO, which value is held by the map $m$ as
that address is now irrelevant. Indeed, $m$ is never read nor modified for
addresses in \MMIO. One could choose to enforce that a particular dummy value is
stored in $m$ for these addresses. Instead, we choose to leave them
unconstrained.

\begin{remark}
  Thomas: Additionally, we might want to disallow in the semantics an execution
  with a \X{PC} register pointing to an $\MMIO$ location (I think this makes
  more sense than reading a random instruction through I/O). Armaël: but it
  makes the semantics of all instructions more complicated.
\end{remark}

\begin{remark} With this presentation, the specification of the
``machine'' is very much decoupled from the model of the devices it might be
communicating with. This is a good thing, I think.

@@

I believe this would be somewhat similar to the semantics of I/O system calls
through a foreign function interfaces as formalized in
CakeML~\cite{cakeml-vstte17io}, and also in Perennial~\cite{perennial-lang}.
%

\end{remark}

\begin{remark} Alix says that the proposed style of operational semantics
(using a trace) is close to the semantics of \texttt{volatile} as in
CompCert---which is also how memory mapped addresses seem to be exposed to C
compilers in practice. So this is probably a good sign. 
\end{remark}

@@

We added a trace as part of the state, so we wish to also expose it as a \SL
assertion. Recall the current definition of the state interpretation:
\[
  \X{state\_interp} \; (r,m) := \X{gen\_heap\_ctx}\; r \ast \X{gen\_heap\_ctx}\; m
\]

@@

The simplest way to account for the trace is to directly expose it in a
monolithic fashion. In that case, the state interpretation additionally holds a
resource for the trace.
\[
  \X{state\_interp} \; (r,m,t) :=
    \X{gen\_heap\_ctx}\; r \ast
    \X{gen\_heap\_ctx}\; (m \setminus \MMIO) \ast
    {\color{BrickRed}\tracefull{t}}
\]

Note that we restrict the usual ``points-to'' assertions to be used only for
non-MMIO addresses. Instead, to assert ownership of the MMIO region, the user
works with assertions of the form $\tracefrag{t'}$. Such an assertion is not
duplicable, and grants full ownership over the trace. In particular, it allows
one to \emph{update} the trace by emitting events, i.e. by performing I/O
operations.

@@

\begin{remark} This is very coarse-grained: either one has the full
ownership for performing I/O and reasoning about it, or one cannot know anything
about it.

@@

One application of this second extension could be the verification of an example
involving ``multiplexed'' I/O, where two separate parts of the code are granted
separate ownership over separate MMIO addresses. These two separate pieces of
code would be verified separately; then, in the end, one could prove that one
gets \emph{some} interleaving of all the events emitted by both components.
\end{remark}

\newcommand{\MMIOag}{\ownGhost{\gamma_{\X{MMIO}}}{\authfull (\X{MMIO})}}

\begin{remark}
  As an alternative to having the set of MMIO addresses (\MMIO) as a global
  parameter, we might want to make it part of the state in the operational
  semantics. In that case, we would need another resource algebra to model the
  MMIO locations, i.e. add do the state interpretation we had before:
%
\[
  \X{state\_interp} \; (r,m,t,\MMIO) := \ldots \ast \MMIOag
\]

(where $\authm(\exm(\X{Trace}))$ is the resource algebra).
\end{remark}

\begin{remark}
In the future, we might be interested in the dynamic allocation of MMIO memory.
The question is what this would mean, though, since the set of available devices
allowing for MMIO access and the concrete buffers they provide will most likely
still be modeled as fixed at runtime.
Rather, this would allow us to take (un)mapped MMIO locations, connected to the different devices, and (un)map them anywhere in main memory.
This would require a different way of modeling MMIO, where we need both a notion
of the total pool of possible MMIO locations, and a description of the currently
mapped locations. We could model this using an authoritative RA.
\end{remark}

@@

One then wants to be able to encapsulate the use of $\OutV$, for instance by
defining object capabilities that enforce some invariants on the values being
printed. In that setting, $\OutV$ is considered as a ``high'' value:
%
\[
  \begin{array}{lcl}
    \liftP \OutV & \eqdef & \FALSE
  \end{array}
\]

The theorem then becomes:
%
\begin{mathpar}
  \mprset{vskip=0.4em}
  \inferrule
  {C \in \textit{AdvCtx} \\
    e \; \X{closed} \\
    \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast \pure{\! P(t)}} \vdash \hoare{\TRUE}{e}{x.\; \textlog{lowval}\; x} \\
    (C[e]);(\emptyset, \textlog{OK}); \emptyset \longrightarrow^* T';(h',g'); t
  }
  {g' = \textlog{OK} \wedge P(t)}
\end{mathpar}

@@

\medskip

To be slightly more general, we could actually allow sharing $\OutV$ in case the
predicate $P$ does not place any constraints on the output, i.e. it is the
$\TRUE$ predicate. Technically, we could hence have the following alternative
definition for $\textlog{lift}$ (although admittedly, it does look a bit
hard-coded):
%
\[
  \begin{array}{lcl}
    \liftP \OutV & \eqdef & \exists \iota \ldotp \knowInv{\iota}{\exists t\ldotp \tracefrag{t} \ast
                            \pure{(\Lam{\_}.\TRUE) \spac t}}
  \end{array}
\]

@@

\subsubsection{Logical relation}

@@

Then, $\VR^{\notMMIO}$ is similar to the $\textlog{lowval}$ predicate of OCPL.
Intuitively, a value in the relation does not directly point to memory-mapped
locations, nor can it gain access to memory-mapped locations indirectly,
similarly to how $\textlog{lowval}$s cannot grant access to high locations
directly. The relation $\ER^{\notMMIO}$ specifies that, if we fill the registers
with values in $\VR^{\notMMIO}$, then the invariants in a private future world
of our starting world will be satisfied at the end of execution.

\begin{remark} As it is defined above, the ``generalized'' value relation
is in fact not very useful for instantiation of $\Psi$ other than $\notMMIO$.
Indeed, even if $\Psi$ \emph{does} allow referring to addresses in the $\MMIO$
region, the value relation does not grant any corresponding resources. The fix
would be to use the generalized trace resources mentioned previously, and grant
ownership for the part of the trace corresponding to the addresses in
$[b,e) \cap \MMIO_{\X{pub}}$ with $\MMIO_{\X{pub}}$ the shareable subset of
addresses in \MMIO.
\end{remark}

\subsubsection{Robust safety theorem}

A first attempt at adapting the OCPL robust safety theorem to the capability
machine setting might look something like the following
{\color{BrickRed}\textbf{(incomplete!)}} statement.

\begin{mathpar}
  \mprset{vskip=0.5em}
  \inferrule
  { \forall a_{i} \in A_{\X{entry}} \ldotp\,
      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\!
         \vdash
         \forall W \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
       \\
     \\
    (r[\overline{r_{a_{i}} := (\X{E}, g, b, e, a_i)}], m, \emptyset) \longrightarrow^*
    (r', m', t)
  }
  {P(t)}
\end{mathpar}

The code that is verified (similar to the expression $e$ in OCPL) is here given
as a set of \emph{entrypoints} (addresses) $A_{\X{entry}}$.
%
The capability $(\X{RX}, g, b, e, a_i)$ then points to the $i$th entry point of
the code we want to gradually verify.

The capability $(\X{E}, g, b, e, a_i)$ then represents a closure for this same piece
of code. Adversarial code only gets access to the closure (as an enter
capability), in order to avoid them from executing or reading arbitrary lines of
code within the trusted module.

Notice the universal quantification over worlds in the precondition;
this needs to be there, since we do not know in what world our code will be
called.

\begin{remark}
  We could simplify the theorem statement by only considering the case of a
  single entrypoint. This is equivalent: a piece of code that wants to expose
  two distinct ``methods'' can implement a single toplevel entrypoint that
  immediately returns two pointers that can then be used to invoke the two
  methods.
  %
  This would make the toplevel theorem simpler, at the cost of somewhat more
  contrieved calling convention between trusted and untrusted code.
\end{remark}

\begin{remark}
The universal quantification on worlds was not explicitly present in the OCPL
formalization.
The reason for that is that (I think so at least, do not quote me
on this - Thomas) they
used the built-in Iris worlds within their definition of weakest precondition,
by leveraging the state interpretation, which is universally quantified over in
the definition of weakest precondition.
As we discussed before, if we do away
with local capabilities, it should be possible to fall back onto Iris' notion of
worlds, which would allow us to remove the universal quantification in the
above definition too because it would be implicitly present in the WP inside $\ER$.

It would be interesting to have a discussion about how to do away with the
worlds if we remove local capabilities, regardless of whether we will in the
end, since that might shed some light on why the worlds look the way they do
right now.
\end{remark}

\begin{remark}
  Note that we could have written
  $\ldots \vdash \forall W \ldotp \; (\VR^{\notMMIO} (\X{E}, \ldots))\; W$ %)
  in the above theorem, instead of using the expression relation. This is
  equivalent, and easily unfolds to the former statement through the definition
  of the value relation. I wrote $\ER$ to place the focus on the execution of
  the driver being safe. I consider this a matter of taste.
\end{remark}

\medskip

The theorem above is \emph{incomplete}. In other words, it is not provable as
is, because it includes no notion of adversarial context (similar to OCPL's
\textit{AdvCtx}), and therefore imposes no restriction on the adversarial code
(i.e., the contents of the $r$ and $m$ other than our verified code).
%
As such, the statement can be trivially falsified by taking an attacker with
direct read or write access to any location in \MMIO, or with direct access to
the trusted code's internal state.

We move to a theorem statement that correctly constrains the adversary in two
steps:
%
\begin{enumerate}
  \item We add a syntactic condition (on the state of memory and registers) to
    the tentative statement of our robust safety theorem, ensuring that the
    adversary code cannot gain access to any non-$\X{E}$ capabilities pointing into
    the driver code or any capabilities pointing into $\MMIO$ directly.
%
  \item Then, we provide an end-to-end statement that assumes the existence of a
    piece of boot code, that gets to run first when the capability machine
    starts up, and makes sure the above condition indeed holds before passing
    control to the adversary.

    The boot code is part of the trusted code base, and must satisfy a given
    specification, as a precondition of the toplevel theorem.

    The particular implementation of the boot code depends on the model of the
    initial state of the machine. If the memory can contain any capability at
    startup, then the boot code has to erase the memory before passing control.
    If we assume that the initial memory cannot contain any capabilities
    whatsoever, then the boot code does not need to do any cleanup.
\end{enumerate}

In this setting, we can subdivide the memory $m$ into 3 relevant subsections;
$m = m_{\MMIO} \uplus m_{\X{driver}} \uplus m_{\X{adv}}$ %\uplus m_{\X{frame}}
with the following meaning:
\begin{itemize}
\item $m_{\MMIO}$ contains the memory-mapped locations
\item $m_{\X{driver}}$ contains the driver code
\item  $m_{\X{adv}}$ contains the adversary's code and data (this will also
  include the boot code in the general setting, since it does not contain any
  inherently dangerous capabilities itself).
\end{itemize}

\newcommand{\nonCap}[1]{\ensuremath{\mathrm{nonCap}(#1)}}

We can then implement the syntactic check as requiring registers and memory to
only contain non-capability values:

\def \MathparLineskip {\lineskiplimit=0.7em\lineskip=0.7em}
\begin{mathpar}
  \nonCap{\X{inl}\; z} \eqdef \TRUE \and
  \nonCap{\X{inr}\; (p,g,b,e,a)} \eqdef \FALSE \and
  \nonCap{\V{reg}} \eqdef \forall r \in \dom(\V{reg}).\; \nonCap{\V{reg}(r)} \and
  \nonCap{\V{mem}} \eqdef \forall l \in \dom(\V{mem}).\; \nonCap{\V{mem}(l)}
\end{mathpar}

@@

Given this definition, we can now formalize the initialization constraints on
registers $r$ and memory $m$, given the region $\MMIO$ and
the driver's address space $[b,e)$ as follows:

\begin{mathpar}
  \mprset{vskip=0.4em}
  \inferrule
  { m = m_{\MMIO} \uplus m_{\X{driver}} \uplus m_{\X{adv}}\\
    \dom( m_{\MMIO} ) = \MMIO \\
    \dom( m_{\X{driver}} ) = [b,e)\\
    \dom( m_{\X{adv}} ) = [b_{\X{adv}},e_{\X{adv}})\\
    \nonCap{\V{r}}\\
    \nonCap{\V{m_{\X{adv}}}} \\
    r[\X{PC}] = (\X{RWX}, \X{G},b_{\X{adv}},e_{\X{adv}},b_{\X{adv}})\\
  }
  {\initOK{[b,e)}}
\end{mathpar}

@@

\begin{mathpar}
  \mprset{vskip=0.5em}
  \inferrule
  { \forall a_{i} \in A_{\X{entry}} \ldotp\,
      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\!
         \vdash
         \forall W \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
         \\
   \initOK{[b,e)} \\
  (r[\overline{r_{a_{i}} := (\X{E}, g, b, e, a_i)}], m, \emptyset) \longrightarrow^* (r', m', t)\\
  }
  {P(t)}
\end{mathpar}

where we have set the PC up to be a sensible capability (pointing to
$b_{\X{adv}}$ without loss of generality).

@@

\begin{remark}
  If useful, the syntactic check can be relaxed to allow any capabilities that
  do not point into the MMIO region or the driver's memory. One would define the
  predicate $\notMapsToR{R}{w}$, that takes a word, memory region or register
  file and checks if it points into the forbidden region $R$ as follows:

\[
  \begin{array}{lcl}
    \notMapsToR{R}{\X{inl}\spac z} & \eqdef
    & \TRUE \\
    %
    \notMapsToR{R}{\X{inr}\spac (p,g,b,e,a)} & \eqdef & R \cap [b,e)\; =
                                                 \emptyset  \\
    %
    \notMapsToR{R}{\V{reg}} & \eqdef & \forall r \in \dom(\V{reg}) \ldotp\spac \notMapsToR{R}{\V{reg}(r)} \\
    %
    \notMapsToR{R}{\V{mem}} & \eqdef & \forall l \in \dom(\V{mem}) \ldotp\spac   \notMapsToR{R}{\V{mem}(l)}
  \end{array}
\]

Then, one would use $\notMapsToR{\MMIO \cup [b,e)}$ in place of $\nonCap{}$.
\end{remark}

Having this theorem in our toolbox, we can now take a closer look at scenario 2
above, where a piece of boot code is used to set the memory up correctly,
obviating the need for any other assumptions than that the boot code behaves
according to some spec. Concretely, we want the boot code to uphold a spec that
allows us to prove the above theorem.
%
(Alternatively, we could also just implement one specific instance of boot code,
and prove that after it finishes executing, all preconditions to apply the above
theorem are met.)

Let $r_0$ and $m_0$ be the respective initial state of registers and memory
immediately after booting. We assume that the following two properties hold,
where $l_\X{boot}$ is some memory location (initially loaded in PC):
%
\begin{mathpar}
  \dom(m_0) = [0,\X{MEM_{MAX}}) \and
  r_0[\X{PC}] = (\X{RWX}, \X{G},0,\X{MEM_{MAX}},l_{\X{boot}})
\end{mathpar}

Depending on the specifics of the operiational semantics, we might have more
information about $m_0$ and $r_0$, that can then be used in the implementation
of the boot code (e.g., $m_0$ might be initialized to only zeroes).

We also assume the boot-code to be initially loaded in memory at address
$l_{\X{boot}}$.
%
Then, the specification for the boot code that we have to prove is stated as
follows:

\newcommand{\bigast}[2]{\underset{#1}\Sep{\!\!\! #2}\;}

\[
  \begin{array}{l}
  \X{bootSpec}(P, g, b, e, A_{\X{entry}}) \eqdef {} \\[0.8em]
    \quad
  \begin{array}{l}
    \left\{
    \begin{array}{l}
      \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, 0, \X{MEM_{MAX}}, l_{\X{boot}}) \ast {} \\[0.5em]
      \bigast{r \in \X{Reg} \setminus \X{PC}}{r \mapsto_{r} \_}
      \ast \bigast{m \in \X{Mem} \setminus \MMIO}{m \mapsto_{a} \_}
      \ast \tracefrag{\emptyset}
    \end{array}
    \right\}
    \\[2em]
    \quad \I{Instr}\; \I{Executable}
    \\[0.5em]
    \left\{
    \begin{array}{l}
      \exists b_{adv}\; e_{adv}\; t.\; \\[0.5em]
      \quad
      \begin{array}{l}
        \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{adv}, e_{adv}, b_{adv}) \ast {} \\[0.7em]
        \bigast{a_{i} \in A_{\X{entry}}}{a_{i} \mapsto_{r} (\X{E},g,b,e,a_i)} \ast
        \bigast{r \in Reg\setminus \{\X{PC}, A_{\X{entry}}\}}{r \mapsto_{r} \X{inl}\; \_} \ast {} \\[1.4em]
        \bigast{m \in [b_{adv}, e_{adv})}{m \mapsto_{a} \X{inl}\; \_} \ast 
        \tracefrag{t} \ast \pure{P(t)}
      \end{array}
    \end{array}
    \right\}
  \end{array}
  \end{array}
\]

Intuitively, establishing this specification ensures that, after running the
boot code, we arrive in a state that satisfies $\initOKo$.

The resulting boot-code-base formulation of the robust safety theorem is now a
corollary of the general robust safety theorem, and has the following statement:

\begin{mathpar}
  \mprset{vskip=0.5em}
  \inferrule
  { \forall a_{i} \in A_{\X{entry}} \ldotp\,
      \knowInv{\iota}{\exists t.\; \tracefrag{t} \ast P(t)}\!
         \vdash
         \forall W \ldotp\;(\ER^{\notMMIO} (\X{RX}, g, b, e, a_i))\spac W
         \\
   \X{bootSpec}(P,g,b,e,A_{\X{entry}}) \\
  (r_0, m_0, \emptyset) \longrightarrow^* (r, m, t)\\
  }
  {P(t)}
\end{mathpar}

Note that to link scenario 2 to scenario 1, we need a slightly adapted version
of the robust safety theorem stated above where the initial trace is allowed to
not be empty, but rather just satisfy $P(t)$, but that should not cause any
problems.
